---
title: 'ECON245 Project'
output: html_notebook
---

# 0. Import libraries 
```{r}
library(dplyr)
library(lubridate)
library(ggplot2)
library(tidyr)
library(glmnet)
library(caret)
library(ROSE)
library(pROC)
library(pscl)
library(vcd)
```

# 1. Import data 
```{r}
data <- read.csv('http://mysmu.edu/faculty/denisleung/101/lun_smk_naga.csv')
```

# 1.1 View data 
```{r}
head(data)
str(data)
```

# 2. Clean data

## 2.1 Check for null and infinite values
```{r}
data%>%summarise_all(~sum(is.na(.))) # no null values 
data%>%summarise_all(~sum(is.infinite(.))) # no infinite values 
```

## 2.2 Check for multicollinearity among categorical variables
```{r}
library(corrplot)
matrix_corr = cor(data) # only find correlations among covariates
corrplot(matrix_corr, method="number", type="upper") # upper diagonal matrix in numbers
```
# Remove numerical variables
```{r}
df <- data[,-c(1,2,14,16,17,18,19,23,24,25,26,27)]
```
# 2.3 EDA

## 2.3 Convert categorical variables into factors
```{r}
df[c('sex', 'un4gy', 'distcat', 'agxcat', 'agecat', 'time', 'dcat', 'scat', 'smkcat', 'smkyrcat', 'smkqyrcat', 'nic')] = lapply(df[c('sex','un4gy', 'distcat', 'agxcat', 'agecat', 'time', 'dcat', 'scat', 'smkcat', 'smkyrcat', 'smkqyrcat', 'nic')],factor)

str(df) # check data
```


### Original categories of 'dcat'
```{r}
ggplot(data = df, aes(x = dcat)) + geom_bar() + labs(title = 'Distribution of weighted lung dose', x = 'dcat', y = 'count')
```

### Recode categories of 'dcat'
```{r}
dcat_sort <- function(x) {
  x <- as.numeric(as.character(x))
  if (x >= 9) {
    return('>= 100 mgy')
  } else if (x >= 4) {
    return('5-100 mgy')
  } else {
    return('0-5 mgy')
  }
}

dcat_order <- c('0-5 mgy', '5-100 mgy', '>= 100 mgy')

df$dcat <- sapply(df$dcat, dcat_sort)

df$dcat <- factor(df$dcat, levels = dcat_order)

ggplot(data = df, aes(x = dcat)) + geom_bar() + labs(title = 'Distribution of weighted lung dose', x = 'dcat', y = 'count')
```
##  Recode 'agxcat'

###  Original categories of 'agxcat'
```{r}
ggplot(data = df, aes(x = agxcat)) + geom_bar() + labs(title = 'Distribution of age at exposure', x = 'agxcat', y = 'count')
```

###  Recode categories of 'agxcat'
```{r}
agxcat_sort <- function(x) {
  x <- as.numeric(as.character(x))
  if (x >= 11) {
    return('Elderly')
  } else if (x >= 5) {
    return('Adult')
  } else if (x >= 3) {
    return('Teenager') 
  } else {
    return('Child')
  }
}

agxcat_order <- c('Child', 'Teenager', 'Adult', 'Elderly')

df$agxcat <- sapply(df$agxcat, agxcat_sort)
63
96

df$agxcat <- factor(df$agxcat, levels = agxcat_order)

ggplot(data = df, aes(x = agxcat)) + geom_bar() + labs(title = 'Distribution of age at exposure', x = 'agxcat', y = 'count')
```
## Recode 'agecat'

### Original categories of 'agecat'
```{r}
ggplot(data = df, aes(x = agecat)) + geom_bar() + labs(title = 'Distribution of attained age', x = 'agecat', y = 'count')
```
### Recode categories of 'agecat'
```{r}
agecat_sort <- function(x) {
  x <- as.numeric(as.character(x))
  if (x >= 15) {
    return('>=70')
  } else if (x >= 13) {
    return('60-70')
  } else if (x >= 11) {
    return('50-60') 
  } else if (x >= 9){
    return('40-50')
  } else {
    return('<=40')
  }
}

agecat_order <- c('<=40', '40-50', '50-60', '60-70', '>=70')

df$agecat <- sapply(df$agecat, agecat_sort)

df$agecat <- factor(df$agecat, levels = agecat_order)

ggplot(data = df, aes(x = agecat)) + geom_bar() + labs(title = 'Distribution of attained age', x = 'agecat', y = 'count')
```

## Recode 'smkcat'

### Original categories of 'smkcat'
```{r}
ggplot(data = df, aes(x = smkcat)) + geom_bar() + labs(title = 'Distribution of smkcat', x = 'smkcat', y = 'count')
```


### Recode categories of 'smkcat's
```{r}
smkcat_sort <- function(x) {
  x <- as.numeric(as.character(x))
  if (x >= 7) {
    return('Heavy Smokers')
  } else if (x >= 5) {
    return('Moderate Smokers')
  } else if (x >= 3) {
    return('Light Smokers') 
  } else if (x == 2){
    return('Non-Smokers')
  } else {
    return('Unknown')
  }
}

smkcat_order <- c('Unknown', 'Non-Smokers', 'Light Smokers', 'Moderate Smokers', 'Heavy Smokers')

df$smkcat <- sapply(df$smkcat, smkcat_sort)

df$smkcat <- factor(df$smkcat, levels = smkcat_order)

ggplot(data = df, aes(x = smkcat)) + geom_bar() + labs(title = 'Distribution of smkcat', x = 'smkcat', y = 'count')
```

## Aggregate all cancer variables into a single 'cancer' variable
```{r}
df <- df %>% mutate(cancer = lung + larynx + othresp)
summary(df$cancer)
```
# Remove 'lung', 'larynx', 'othresp' columns
```{r}
df <- df[,-c(13,14,15)]
str(df)
```

## 3 Identifying Statistically Significant Cancer Risk Factors
```{r}
# count
table(df$cancer)

# proportion of non-cancer vs cancer cases 
df %>%
  group_by(cancer) %>%
  summarise(cnt = n()) %>%
  ungroup() %>%
  mutate(proportion = cnt / sum(cnt)) %>%
  ggplot(aes(x = factor(cancer), y = proportion, fill = factor(cancer))) +
  geom_bar(stat = 'identity') +
  geom_text(aes(label = scales::percent(proportion), y = proportion + 0.01), vjust = -0.5) +
  ggtitle('Proportion of Cancer Counts') +
  xlab('Cancer Status') +
  ylab('Proportion')
```

## 5.3 Proportion of 'cancer' counts by 'sex'
```{r}
df %>%
  group_by(cancer, sex) %>%
  summarise(cnt = n()) %>%
  ungroup() %>%
  mutate(proportion = cnt / sum(cnt)) %>%
  
ggplot(aes(x = factor(cancer), y = proportion, fill = sex)) +
  geom_bar(position = 'stack', stat = 'identity') +
  geom_text(aes(label = scales::percent(proportion), y = proportion + 0.01), vjust = -0.5) +
  labs(x = 'Cancer', y = 'Number of Cases', title = 'Stacked Bar Plot of Cancer Cases by Sex')
```


## 5.4 Distribution of 'cancer' over 'time'
```{r}
df%>%
  group_by(time, cancer)%>%
  summarise(cnt=n(), .groups='drop')%>%
  filter(cancer == 1)%>%
  ggplot(aes(x=time, y = cnt)) + geom_bar(stat='identity') + ggtitle('Distribution of Cancer Cases over Time')
```
# 3.1 Chi Squared test
## 3.1.1 Smoking (scat) chi-squared test
Since the p-value is < 2.2e-16 < 0.05, there is sufficient evidence to reject the null hypothesis. Therefore, there is a significant relationship between smoking status and cancer occurrence.
```{r}
table(df$cancer, df$scat)

# Visualise the proportion of each category 
mosaic(~ cancer + scat,
  direction = c("v", "h"),
  data = df,
  shade = TRUE
)

# Conduct chi-sq test 
scat_test <- chisq.test(table(df$cancer, df$scat))
scat_test
```
## 3.1.1 Two-sample Poisson test on the number of cancer cases, stratified by smoking status (scat)

# Two-pop Poisson for Smokers and Non-smokers
```{r}
# Get population that smokes vs those who do not 
no_smoke_df = filter(df, scat %in% c(2, 3))
smoke_df = filter(df, scat==4)

# two-sample poisson test
poisson.test(c(sum(no_smoke_df$cancer), sum(smoke_df$cancer)), c(nrow(no_smoke_df), nrow(smoke_df)),
alternative = "less", conf.level = 0.95)

c(sum(no_smoke_df$cancer))/nrow(no_smoke_df)
c(sum(smoke_df$cancer))/nrow(smoke_df)
```

## 3.1.2 Radiation (dcat) chi-squared test

```{r}
# Contingency table between cancer and dcat
table(df$cancer, df$dcat)

# Visualise the proportion of each category 
mosaic(~ cancer + dcat,
  direction = c("v", "h"),
  data = df,
  shade = TRUE
)

# Conduct chi-sq test 
dcat_test <- chisq.test(table(df$cancer, df$dcat))
dcat_test
```
## 3.1.3 Sex (sex) chi-squared test

```{r}
# Contingency table between cancer and sex
table(df$cancer, df$sex)

# Visualise the proportion of each category 
mosaic(~ cancer + sex,
  direction = c("v", "h"),
  data = df,
  shade = TRUE
)

# Conduct chi-sq test 
sex_test <- chisq.test(table(df$cancer, df$sex))
sex_test
```
# Two-pop Poisson for Male and Female
```{r}
# Get population of males and females
male_df = filter(df, sex==1)
female_df = filter(df, sex==2)

# two-sample poisson test
poisson.test(c(sum(male_df$cancer), sum(female_df$cancer)), c(nrow(male_df), nrow(female_df)),
alternative = "greater", conf.level = 0.95)

c(sum(male_df$cancer))/nrow(male_df)
c(sum(female_df$cancer))/nrow(female_df)
```

## 3.1.4 Age at exposure (agxcat) chi-squared test
  
```{r}
# Contingency table between cancer and agxcat
table(df$cancer, df$agxcat)

# Visualise the proportion of each category 
mosaic(~ cancer + agxcat,
  direction = c("v", "h"),
  data = df,
  shade = TRUE
)

# Conduct chi-sq test 
agxcat_test <- chisq.test(table(df$cancer, df$agxcat))
agxcat_test
```
# Two sample poisson test to compare the mean between each age group of exposure
```{r}
elderly_df <- filter(df, agxcat=="Elderly")
adult_df <- filter(df, agxcat=="Adult")
teenager_df <- filter(df, agxcat=="Teenager")
child_df <- filter(df, agxcat=="Child")

# child vs adult
poisson.test(c(sum(child_df$cancer), sum(adult_df$cancer)), c(nrow(child_df), nrow(adult_df)),
alternative = "less", conf.level = 0.95)

# teenager vs adult 
poisson.test(c(sum(teenager_df$cancer), sum(adult_df$cancer)), c(nrow(teenager_df), nrow(adult_df)),
alternative = "less", conf.level = 0.95)

# adult vs elderly 
poisson.test(c(sum(adult_df$cancer), sum(elderly_df$cancer)), c(nrow(adult_df), nrow(elderly_df)),
alternative = "less", conf.level = 0.95)
```
# 3.1.5 Not in city (NIC) chi-squared test
```{r}
# Visualise the proportion of each category 
mosaic(~ cancer + nic,
  direction = c("v", "h"),
  data = df,
  shade = TRUE
)

# Conduct chi-sq test 
nic_test <- chisq.test(table(df$cancer, df$nic))
nic_test
```
# Two-pop Poisson for NIC
```{r}
# Get population that smokes vs those who do not 
nic_df = filter(df, nic==1) #not in the city at time of bombing
ic_df = filter(df, nic==0) #in the city at time of bombing

# two-sample poisson test
poisson.test(c(sum(nic_df$cancer), sum(ic_df$cancer)), c(nrow(nic_df), nrow(ic_df)),
alternative = "less", conf.level = 0.95)

c(sum(nic_df$cancer))/nrow(nic_df)
c(sum(ic_df$cancer))/nrow(ic_df)
```
## 3.2 Stepwise regression (backward elimination)
```{r}
df_stepwise <- read.csv("http://mysmu.edu/faculty/denisleung/101/lun_smk_naga.csv")

# running only numerical variables 
dfstepwise <- df_stepwise %>% dplyr :: select(-c(X, city, gdist, agex, age, nic, year, d10lun, smkamt, smkyrs, smkqyrs, pyr, upyr))

# adding cancer variables 
dfstepwise <- dfstepwise %>% mutate(cancer = lung + larynx + othresp) %>% dplyr::select(-c(lung, larynx, othresp))

# define model with all predictors
all <- glm(cancer ~ sex+un4gy+distcat+agxcat+agecat+time+dcat+scat+smkcat+smkyrcat+smkqyrcat+subjects+offset(log(pyr92)), data=dfstepwise, family='poisson')

# perform backward stepwise regression
backward <- step(all, direction='backward', scope=formula(all), trace=0)

# view results of backward stepwise regression
backward$anova

# view final model to see significant variables 
backward$coefficients
```
# 4. Models

# 4.1 Poisson Regression
# Model 1
```{r}
out1 = glm(cancer~sex+agecat+time+dcat+scat+offset(log(pyr92)), family='poisson', data=df)
summary(out1)
``` 
# Model 2
```{r}
out2 = glm(cancer~sex+agecat+time+dcat+scat+scat:dcat+offset(log(pyr92)), family='poisson', data=df)
summary(out2)
``` 
# Model 3
```{r}
out3 = glm(cancer~sex+agecat+time+dcat+scat+sex:dcat+offset(log(pyr92)), family='poisson', data=df)
summary(out3)
``` 
# Model 4
```{r}
out4 = glm(cancer~sex+agecat+time+dcat+scat+agecat:dcat+offset(log(pyr92)), family='poisson', data=df)
summary(out4)
``` 
# Model 5
```{r}
out5 = glm(cancer~sex+agecat+time+dcat+scat+time:dcat+offset(log(pyr92)), family='poisson', data=df)
summary(out5)
``` 
# Model 6
```{r}
out6 = glm(cancer~sex+agecat+time+dcat+scat+agecat:sex+offset(log(pyr92)), family='poisson', data=df)
summary(out6)
``` 
# 4.2 Hurdle Regression

# Model 7
```{r}
out7 = hurdle(cancer~sex+dcat+scat+agecat+time+offset(log(pyr92)), data=df)
summary(out7)
``` 

# 4.3 Zero inflated Regression

# Model 8
```{r}
out8 = zeroinfl(cancer~sex+dcat+scat+agecat+time+offset(log(pyr92)), data=df, dist='poisson', link = 'logit')
summary(out8)
``` 
# 4.4 Logistic Regression

# Model 9
```{r}
df <- df %>%
  mutate(cancer_binary = ifelse(cancer > 0, 1, 0))
out9 = glm(cancer_binary~sex+dcat+scat+agecat+time, data=df, family='binomial')
summary(out9)
``` 
# 5 Results
# 5.2 AIC and BIC scores
```{r}
AIC(out1, out2, out3, out4, out5, out6, out7, out8, out9)
```
```{r}
BIC(out1, out2, out3, out4, out5, out6, out7, out8, out9)
```



